---
title: "Introduction to R"
author: "Priit Adler"
date: "2025-04-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction (Rstudio / R)

You can find this and the data files from ELIXIREstonia github:
<https://github.com/ELIXIREstonia/2025-04-24-R-visualisation>

R is a popular open-source programming language and environment for statistical
computing and graphics.

RStudio is an integrated development environment (IDE) tailored for efficient
and user-friendly R programming. Once launched, the RStudio UI presents itself
in a segmented manner, primarily divided into four main panes.

-   Text editor

-   Console / Terminal / etc...

-   Environment / History / Connections / Git

-   Files / **Plots** / Packages / **Help** / etc...

see: <https://rstudio.github.io/cheatsheets/html/rstudio-ide.html> for more
detailed overview.

This is the second part of a two-part curriculum designed to introduce R and
ggplot2. If you are unfamiliar with the basic syntax, data types, and the
tidyverse in R, then you should first check out Part One at:
<https://github.com/ELIXIREstonia/2025-04-23-R-basic>.

## Loading the Tidyverse

```{r message=FALSE}
# Install tidyverse if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

# Load the tidyverse
library(tidyverse)
```

## Reading Data from a File

R supports a wide variety of file formats for reading and writing data. Some
common formats include:

-   CSV (Comma-Separated Values)
-   TSV (Tab-Separated Values)
-   Excel (XLS, XLSX)
-   .. and few others

The `tidyverse` packages, such as `readr` and `readxl`, provide functions for
reading data from many of these formats. In this workshop, we'll focus on
reading data from CSV, TSV.

## About the data

In examples we'll be using a
[Kaggle](https://www.kaggle.com/datasets/steveahn/memory-test-on-drugged-islanders-data?resource=download)
"**Memory Test on Drugged Islanders Data**" dataset.

Dataset description

:   An experiment on the effects of anti-anxiety medicine on memory recall when
    being primed with happy or sad memories. The participants were done on novel
    Islanders whom mimic real-life humans in response to external factors.

    Drugs of interest (known-as) [Dosage 1, 2, 3]:

    A - Alprazolam (Xanax, Long-term) [1mg/3mg/5mg]

    T - Triazolam (Halcion, Short-term) [0.25mg/0.5mg/0.75mg]

    S- Sugar Tablet (Placebo) [1 tab/2tabs/3tabs]

The easiest way to read in the data is to use `read_csv` command:

```{r}
Isd <- read_csv('data/Islander_data.csv')

Isd
```

# High-Level Overview of the Data with Tidyverse Pipes

Tidyverse pipes, represented by the `%>%` \<ctrl+shift+M\> symbol, allow you to
chain together multiple functions in a clear and readable manner. Pipes take the
output of one function and use it as the input for the next function, making it
easy to follow the sequence of data transformations.

In the following example, we'll use pipes to perform a series of operations on
our dataset: 1. Group the data by a specific category (`group_by`) 2. Calculate
summary statistics for each group (`summarize`) 3. Sort the resulting summary by
a specific statistic (`arrange`)

```{r include=FALSE}
# Create a dummy dataset
set.seed(42)
data <- tibble(
  YourCategory = factor(sample(LETTERS[1:5], 100, replace = TRUE)),
  YourVariable = rnorm(100, mean = 50, sd = 10),
  YourOtherVariable = rnorm(100, mean = 20, sd = 2)
)
head(data)
```

```{r}
# Summarize the data
data_summary <- data %>% 
  group_by(YourCategory) %>%
  summarize(
    mean_value = mean(YourVariable, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(desc(mean_value))

head(data_summary)
```

In the above code, replace `YourCategory` and `YourVariable` with the
appropriate column names from your dataset. This will give a high-level overview
of the data, summarizing it by the specified category and calculating the mean
of the specified variable. The use of pipes makes it easy to understand the
sequence of transformations applied to the data.

```{r}
# New column using mutate and case_when statement
# note that we are overwriting (updating) the original data
data <- data %>%
  mutate(
    HalfOfOtherVariable = YourOtherVariable / 2, # we can add another column based on existing columns
    YourGroup = case_when(
      YourCategory %in% c("A", "B") ~ "Group 1", # combine A and B into Group 1
      YourCategory %in% c("C", "D") ~ "Group 2", # combine C and D into Group 2
      TRUE ~ "Other"                             # case_when needs fallback default value, 
                                                 # if none of the above match
    ),
    NewVariable = case_when(
      YourVariable > 60 ~ YourVariable * 2,  # Double values above 60
      YourVariable < 40 ~ YourVariable / 2,  # Halve values below 40
      TRUE ~ YourVariable                    # Keep as is otherwise
    )
  )
head(data)
```

Rstudio developers host number of useful
[cheatsheets](https://posit.co/resources/cheatsheets/), take a look. For example
[data transformations with
dplyr](https://rstudio.github.io/cheatsheets/data-transformation.pdf) (dplyr is
part of tidyverse)

### Task:

**Task: Analyzing Effects of Drugs on Memory**

Using the "Memory Test on Drugged Islanders Data" dataset, your task is to
analyze the effects of the different drugs and their dosages on memory recall.

1.  **Age Group Column**:

    -   Create a new column in the dataset called **`age_group`**. Categorize
        participants as "Young" if they are below 30, "Middle-aged" if they are
        between 30 and 50, and "Senior" if they are above 50. (*hint:*
        `case_when`)

```{r}



```

# Plotting with ggplot2

The ggplot2 package in R follows a modular paradigm based on the "Grammar of
Graphics." This modular approach allows users to build complex plots by
combining simple components or layers. Each layer represents a specific element
of the plot, such as data, aesthetics, geoms, scales, and themes.

1.  **Data**: This is the foundation of any plot. You specify the dataset you
    want to visualize.

2.  **Aesthetics**: Aesthetics are the visual properties of the plot, such as x
    and y position, color, size, shape, and transparency. You map variables in
    the dataset to these aesthetics, creating a relationship between the data
    and the plot elements.

3.  **Geoms**: Geoms (short for geometric objects) are the actual plot elements,
    such as points, lines, and bars. Different geoms represent different types
    of plots, like scatter plots, line plots, or bar plots. You can add multiple
    geoms to a single plot to create complex visualizations.

4.  **Scales**: Scales control how data values are mapped to aesthetic
    properties. They define the transformation and mapping of data values to
    visual properties, such as colors, sizes, or shapes. You can adjust scales
    to customize the appearance of the plot.

5.  **Themes**: Themes control the non-data aspects of the plot, such as the
    background, gridlines, axis labels, and legend. You can customize the plot's
    appearance by changing its theme.

In the ggplot2 modular paradigm, you start by specifying the data and the
aesthetics, then add geoms, scales, and themes as needed. This layer-by-layer
approach allows you to create a wide range of plots by combining and customizing
these components.

Here's an example that demonstrates the modular paradigm:

```{r}
# Define data and aesthetics
plot <- ggplot(data = iris %>% filter(Petal.Length > 4), aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point() + 
  scale_color_continuous(low = "blue", high = "red") + 
  theme_minimal()

# Display the final plot
plot
```

In this example, we first define the data and aesthetics, then add a point geom,
adjust the color scale, and finally apply a minimal theme. The result is a
scatter plot that shows the relationship between car weight and miles per
gallon, with points colored according to horsepower.

## Aesthetics in ggplot2

Aesthetics are the visual properties of the elements in a plot. They help convey
the underlying patterns and relationships in the data. In ggplot2, you map
variables from your dataset to aesthetics to create a relationship between the
data and the plot elements. Common aesthetics include `x` and `y` position,
`color`, `size`, `shape`, `group`, and `transparency`.

### Mapping Aesthetics

Let's create a scatter plot with the `age` variable mapped to the x-axis, the
`Diff` variable mapped to the y-axis, and the `Drug` variable mapped to the
color aesthetics:

```{r}
# Create a scatter plot with aesthetics mapped to variables
scatter_plot <- ggplot(Isd, aes(x = age, y = Diff, color = factor(Drug))) +
  geom_point()

scatter_plot
```

### Modifying Aesthetic Properties and Overwriting in Geoms

You can modify aesthetic properties directly within a geom. This allows you to
make specific adjustments to the appearance of individual plot elements. Let's
create a scatter plot with larger points and a custom transparency:

```{r}
# Modify the size and transparency of points within the geom
scatter_plot_independent_aes <- ggplot(Isd, aes(x = age, y = Diff)) +
  geom_point(size = 3, alpha = 0.3, color='blue')

scatter_plot_independent_aes

scatter_plot_data_related_aes <- ggplot(Isd, aes(x = age, y = Diff)) +
  geom_point(size = 3, alpha = 0.5, aes(color = factor(Drug)))

scatter_plot_data_related_aes

```

In this example, we have increased the size of the points and made them
semi-transparent by setting the `size` and `alpha` arguments within the
`geom_point()` function, respectively. By modifying aesthetics directly within a
geom, you can overwrite the initial mappings created by the `aes()` function and
gain more control over the appearance of your plot elements.

## Exploring Different Geom Selections

In this section, we will demonstrate various geom selections using simple dummy
data. Each geom represents a specific type of plot, and their applicability
depends on the type of data (numeric vs. categorical) and the relationship you
want to visualize.

### Scatterplot (geom_point)

A scatterplot displays the relationship between two numeric variables by
plotting points at their respective x and y coordinates. It's useful for
visualizing trends, patterns, or outliers in the data.

```{r fig.width=6, fig.height=6}
# Use Islander dataset as an example
head(Isd)

# Create a scatterplot of memory score before vs memory score after taking the drug
scatter_plot <- ggplot(Isd, aes(x = Mem_Score_Before, y = Mem_Score_After)) +
  geom_point()

scatter_plot
```

### Line Plot (geom_line)

A line plot connects data points with lines to visualize the relationship
between two numeric variables. It's useful for showing trends over time or any
continuous variable.

```{r}
# Use the pressure dataset as an example
head(pressure)

# Create a line plot of temperature vs. pressure
line_plot <- ggplot(pressure, aes(x = temperature, y = pressure)) +
  geom_line()

line_plot
```

### Bar Plot (geom_bar) and Column Plot (geom_col)

A bar plot displays the frequency or count of categorical data, while a column
plot displays the value of a numeric variable for each category. Both are useful
for visualizing relationships between categorical and numeric variables.

```{r}
# Use Islander dataset as an example
head(Isd)

# on categorical data, count how many instances of each appears in the data
bar_plot <- ggplot(Isd, aes(x = Drug)) +
  geom_bar()

bar_plot

# on numerical data (y-axis) sum up all values, relative to x axis categories
column_plot <- ggplot(Isd, aes(x = Drug, y = Diff)) +
  geom_col()

column_plot
```

### Histogram (geom_histogram)

A histogram groups numeric data into bins and displays the frequency of
observations in each bin. It's useful for visualizing the distribution of a
numeric variable.

```{r}
# Create a histogram of the mpg variable from the mtcars dataset
histogram_plot <- ggplot(Isd, aes(x = age)) +
  geom_histogram(binwidth = 2)

histogram_plot
```

### Box Plot (geom_boxplot)

A box plot displays the distribution of a numeric variable across different
categories. It's useful for comparing distributions and identifying outliers
within categorical groups.

```{r}
# Create a box plot of price by cut for the diamonds dataset
box_plot <- ggplot(Isd, aes(x = Drug, y = Diff)) +
  geom_boxplot()

box_plot
```

These are just a few examples of the many `geoms` available in ggplot2. By
selecting the appropriate geom for your data, you can create informative and
visually appealing plots that effectively communicate the relationships within
your dataset.

## Position Adjustments and Adding Multiple Layers

### Jittering Points (position = "jitter")

In scatter plots, data points can sometimes overlap, making it difficult to
discern individual observations. To address this issue, you can apply a position
adjustment, such as `jitter`, which slightly moves the points in a random
direction to reduce overlap.

```{r}

jittered_plot <- ggplot(Isd, aes(x = Drug, y = Diff)) +
  geom_point(position = position_jitter(width = 0.4), alpha = 0.5) +
  geom_boxplot(aes(fill = Drug), alpha = 0.2, outlier.shape = NA) +
  theme_minimal()

jittered_plot
```

In this example, the points are `jittered` to reduce overlap, making it easier
to see the distribution of observations within each `cut` category.

### Adding Multiple Layers (Geoms)

You can add multiple `geoms` to a single plot to create more complex
visualizations. For example, you can combine a scatter plot with a smoothed line
to show the overall trend, and add text labels to annotate specific data points.

```{r}
# Use the mtcars dataset as an example
head(mtcars)

# Create a scatter plot of mpg (miles per gallon) vs. wt (weight) with a smoothed line
scatter_plot_smooth <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "loess", se = FALSE, linetype = "dashed", color = "blue") +
  theme_minimal()

scatter_plot_smooth

# Annotate specific data points with text labels
scatter_plot_annotate <- scatter_plot_smooth +
  geom_text(data = subset(mtcars, mpg > 30 | wt > 5), 
            size = 3, hjust = 0.25, vjust = 1.5, 
            mapping = aes(label=rownames(subset(mtcars, mpg > 30 | wt > 5)))) +
  annotate(geom = 'text', x = 5, y = 30, label="Text that I can add here", color="blue")

scatter_plot_annotate
```

In this example, we first created a scatter plot of `mpg` vs. `wt` and added a
smoothed line using `geom_smooth`. Then, we annotated specific data points with
text labels using `geom_text`. By combining multiple `geoms`, you can create
more informative and visually appealing plots.

### Task:

Use the cell below to repeat the steps we have done so far. Generate a plot
where you show `Islander_data` data by `Drug` and `Diff`. Use `geom_violin`,
what does it show and how to interpret it? Add `geom_point` to there as well,
make it very light (transparent), use position `jitter`. Find more options from
[data-transformations](https://rstudio.github.io/cheatsheets/data-transformation.pdf)
cheatsheet.

```{r fig.width=8, fig.height=6}



```

## Scaling Numeric and Discrete Variables

Scaling variables allows you to transform their range, making it easier to
visualize data that spans multiple orders of magnitude or to compare multiple
variables with different units.

### Scaling Numeric Variables

For numeric variables, you can use `scale_x_continuous()` and
`scale_y_continuous()` to adjust the scales of the `x` and `y` axes. The
`mtcars` dataset, which is built into R, provides a great example for
demonstrating this.

Example: Log Transformation of Horsepower Let's create a scatter plot of miles
per gallon (mpg) vs. horsepower (hp) and scale the `y` axis using log
transformations:

```{r}

# Scatter plot with log-scaled y-axis using mtcars data
scatter_plot_scaled <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(trans = "log10")

scatter_plot_scaled
```

In this example, using log transformations to scale the horsepower helps reveal
patterns in the relationship between fuel efficiency (mpg) and engine power (hp)
that might not be apparent with the original scales.

### Scaling Discrete Variables

For discrete variables, you can use `scale_x_discrete()` and
`scale_y_discrete()` to modify the order or appearance of the categories. With
`mtcars`, we can explore how the number of cylinders (cyl) in a car affects its
fuel efficiency (mpg).

Example: Reordering Cylinder Categories by Median MPG Let's create a box plot of
the number of cylinders vs. mpg and reorder the cylinder categories based on
their median mpg:

```{r}
# Calculate the median mpg for each cylinder category
cyl_median <- mtcars %>%
  group_by(cyl) %>%
  summarize(median_mpg = median(mpg, na.rm = TRUE)) %>%
  arrange(median_mpg)

# Create a box plot with reordered cylinder categories
box_plot_scaled <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  scale_x_discrete(limits = factor(cyl_median$cyl)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Number of cylinders")

box_plot_scaled
```

In this example, reordering the number of cylinders based on the median miles
per gallon helps highlight differences in fuel efficiency across different
engine sizes. This makes it easier to compare the fuel efficiency of cars with
varying numbers of cylinders.

Scaling variables can enhance the readability of your plots and reveal hidden
patterns in your data. By applying appropriate transformations to numeric and
discrete variables, you can create more effective visualizations.

## Color Scaling

Color scaling can be used to visualize an additional variable in a plot, adding
an extra dimension of information. In this revised example, we'll create a box
plot of the number of cylinders (**`cyl`**) vs. miles per gallon (**`mpg`**),
with the color of the points representing engine displacement (**`disp`**).

```{r}
# Create a box plot with reordered cylinder categories and color scaling
box_plot_color_scaled <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +  # fill based on disp
  geom_jitter(aes(color = disp), width = 0.3) +  # jitter points colored by disp
  scale_x_discrete(limits = factor(cyl_median$cyl)) +
  scale_color_gradient(low = "blue", high = "red") +  # gradient color for jitter points
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

box_plot_color_scaled
```

In this example, **`scale_fill_gradient`** and **`scale_color_gradient`** are
used to adjust the color scales based on the engine displacement (**`disp`**).
The box plots and jitter points are colored from blue (lower displacement) to
red (higher displacement), providing an extra layer of information on top of the
relationship between the number of cylinders and mpg.

By applying color scaling, you create richer visualizations that display more
information and enhance the understanding of your data. This technique is
particularly useful in exploring and presenting complex relationships in a
dataset.

## Refactoring and Reordering Factorial Data

Factor variables are categorical variables that can take a limited number of
distinct values. In ggplot2, you can refactor and reorder factor variables to
enhance your visualizations.

### Refactoring

Refactoring involves changing the levels of a factor variable. You can use the
`forcats` package (part of the tidyverse) to refactor variables. The
`fct_recode()` function can be used to recode the levels of a factor. See also
[cheatsheet](https://rstudio.github.io/cheatsheets/factors.pdf).

```{r}
# Load the required packages
# library(forcats)

# # Recode the Drug variable
Isd_refactored <- Isd %>%
  mutate(Drug_name = fct_recode(Drug,
                                "Alprazolam" = "A",
                                "Triazolam" = "T",
                                "Sugar" = "S"))

head(Isd_refactored)
```

In this example, we refactored the `Drug` variable by changing the levels from
letters to actual drug names.

### Reordering

Reordering involves changing the order of the levels of a factor variable. You
can use the `fct_reorder()` function to reorder the levels of a factor based on
the values of another variable.

```{r}
# Create a new column with reordered genre based on the median intgross
Isd_reordered <- Isd_refactored %>%
  mutate(Drug_reordered = fct_reorder(Drug_name, Diff, .fun = max, .desc = TRUE))

head(Isd_reordered)
```

In this example, we reordered the levels of the `Drug` variable based on the
median value of `Diff` for each genre. This can be useful for creating plots
where the categories are ordered meaningfully.

Now let's create a bar plot of the reordered Drug_reordered with the refactored
Drug_name variable:

```{r}
# Create a bar plot of the reordered Drug and Happy_Sad_group variables
bar_plot_reordered <- ggplot(Isd_reordered, aes(x = Drug_reordered, fill = Happy_Sad_group)) +
  geom_bar(position = "dodge") +
  labs(x = "Drug (reordered)", fill = "Happy/Sad Group") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

bar_plot_reordered
```

In this plot, we used the refactored and reordered `Drug` variable to create a
more informative visualization. By refactoring and reordering factor variables,
you can enhance the readability and effectiveness of your plots.

## Enhancing Plots with Titles, Axis Labels, Limits, and Themes

Adding titles, axis labels, limits, and customizing the theme can make your
plots more informative and visually appealing. In this section, we will
demonstrate how to enhance a plot using these features.

Let's create a scatter plot of **`Sepal.Length`** vs **`Sepal.Width`** and use
the **`Petal.Length`** column for color scaling in the **`iris`** dataset:

```{r}

# Create a scatter plot of Sepal.Length vs Sepal.Width and use Petal.Length for color scaling
scatter_plot_example <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Petal.Length)) +
  geom_point(alpha = 0.7) +
  scale_color_continuous(low = "blue", high = "red") +
  theme_minimal()

scatter_plot_example
```

### Adding Title and Axis Labels

You can add a main title, subtitle, and axis labels using the `labs()` function:

```{r}
# Add a main title, subtitle, and axis labels
scatter_plot_labeled <- scatter_plot_example +
  labs(
    title = "Sepal Length vs. Sepal Width",
    subtitle = "Colored by Petal Length",
    x = "Sepal Length",
    y = "Sepal Width",
    color = "Petal Length"
  )

scatter_plot_labeled
```

### Setting Axis Limits

To set the limits for the x and y axes, you can use the `xlim()` and `ylim()`
functions:

```{r}
# Set the limits for the x and y axes
scatter_plot_limited <- scatter_plot_labeled +
  xlim(4, 8) +
  ylim(2, 4) + 
  geom_point(mapping = aes(size = Petal.Width), alpha=0.4) +
  labs(size = 'Petal Width')

scatter_plot_limited
```

### Customizing the Theme

You can customize the appearance of your plot by modifying the theme. The
`theme()` function allows you to change various aspects of the plot, such as
text size, font, background colors, and grid lines:

```{r}
scatter_plot_custom_theme <- scatter_plot_limited +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "gray", linetype = "dashed", linewidth = 0.5),
    panel.grid.minor = element_line(color = "gray", linetype = "dotted", linewidth = 0.25)
  )

scatter_plot_custom_theme
```

By adding titles, axis labels, limits, and customizing the theme, you can create
more informative and visually appealing plots that effectively communicate your
data's story. See
[cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf) for
additional themes.

## Exporting and Saving Plots with ggsave

`ggsave()` is a convenient function for saving your ggplot2 plots in various
file formats, such as PNG, PDF, SVG, or TIFF. Using `ggsave()`, you can easily
save high-quality versions of your plots for use in reports, presentations, or
publications.

To save this plot as a high-quality PNG image, you can use the `ggsave()`
function:

```{r}
# Save the plot as a PNG file
ggsave(
  filename = "scatter_plot_example.png",
  plot = scatter_plot_example,
  width = 8,
  height = 5,
  dpi = 300
)
```

In this example, we saved the `scatter_plot_example` plot as a PNG file with a
width of 8 inches, a height of 5 inches, and a resolution of 300 dots per inch
(DPI). You can adjust the `width`, `height`, and `dpi` parameters to control the
size and quality of the saved image. Here `plot =` argument may be optional --
if plot is not assigned it will by default take the last active plot.

To save the plot in a different file format, you can change the file extension
in the `filename` parameter. For example, to save the plot as a PDF, you can
use:

```{r}
# Save the plot as a PDF file
ggsave(
  filename = "scatter_plot_custom_theme.pdf",
  plot = scatter_plot_custom_theme,
  width = 8,
  height = 5
)
```

By using `ggsave()`, you can easily export and save your ggplot2 plots in a
variety of file formats to share or include in your documents.

## Combining Multiple Plots with patchwork

The `patchwork` package allows you to easily combine multiple ggplot2 plots into
a single layout. This can be useful for comparing different visualizations side
by side or creating more complex visualizations.

First, let's create some example plots using the `mtcars` dataset:

```{r}
# Load the required packages
library(ggplot2)
library(patchwork)

# Create a scatter plot of mpg vs. wt
scatter_plot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = cyl)) +
  labs(title = "Miles per Gallon vs. Weight", x = "Weight (1000 lbs)", y = "Miles per Gallon", color = "Cylinders") +
  theme_minimal()

# Create a bar plot of the number of cars per number of cylinders
bar_plot <- ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(aes(fill = factor(cyl))) +
  labs(title = "Number of Cars per Number of Cylinders", x = "Number of Cylinders", y = "Count", fill = "Cylinders") +
  theme_minimal()

# Create a box plot of mpg per number of cylinders
box_plot <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(aes(fill = factor(cyl))) +
  labs(title = "Miles per Gallon per Number of Cylinders", x = "Number of Cylinders", y = "Miles per Gallon", fill = "Cylinders") +
  theme_minimal() # + theme(legend.position = "None")
```

Now, let's use the `patchwork` package to combine these plots:

```{r}
# Combine the plots using patchwork
combined_plot <- scatter_plot + bar_plot + box_plot + plot_layout(ncol = 1)
combined_plot
```

In this example, we combined the three plots into a single column layout. You
can adjust the layout by changing the `ncol` and `nrow` parameters in the
`plot_layout()` function.

To collect the legends and add a global title, subtitle, and caption, you can
use the `plot_annotation()` function:

```{r fig.height=8, fig.width=10}
# Collect legends and add global title, subtitle, and caption
combined_plot_annotated <- scatter_plot / (bar_plot + box_plot) +
  plot_annotation(
    title = "Exploring the mtcars Dataset",
    subtitle = "Scatter plot, bar plot, and box plot",
    caption = "Data source: mtcars",
    tag_levels = 'A'
  ) +
  plot_layout(guides = "collect")

combined_plot_annotated
```

By using the `patchwork` package, you can combine multiple ggplot2 plots into a
single layout, making it easier to compare and present your visualizations.

### Task:

Now it's your time to shine. Let your imagination fly and explore the Islander
example dataset. Come up with your own visualisation. Use as many plots, as many
colours and themes as you see fit. Start building it gradually, this makes
finding and identifying errors easier. You can even take your own data and
explore it. But try to get to a publishing ready image (*sans* figure legend).

Explore more options from various
[cheatsheets](https://posit.co/resources/cheatsheets/).

```{r}



```

#### Bonus

Explore
['esquisse'](https://cran.r-project.org/web/packages/esquisse/vignettes/get-started.html)
package for interactive plot building. It starts a shiny app to allow
drag-and-drop graph building.
